Supporting Dynamic Data Partitioning in Distributed Memory Architectures with Logical Regions

### 2012-11-04 doc/types-pldi2013/conclusion.tex
spell checking corrections and space saving measures

%Modern architectures have dramatically increased in complexity in recent years.  To program this class
%of machines, new programming systems will be required that are capable about reasoning about the structure
%of data including aliasing.  
We have presented static and dymamic semantics for the Legion programming system,
showing how a combination of static and dynamic checks can be used to soundly and efficiently schedule
Legion programs in parallel, even in the presence of dynamically partitioned and aliased data.  We have
also given a novel compositional parallel semantics, permiting a precise treatment of relaxed coherence modes;
in particular we have shown the Legion design is sound even with relaxed coherence.

>>>>

%Modern architectures have dramatically increased in complexity in recent years.  To program this class
%of machines, new programming systems will be required that are capable about reasoning about the structure
%of data including aliasing.  
We have presented static and dynamic semantics for the Legion programming system,
showing how a combination of static and dynamic checks can be used to soundly and efficiently schedule
Legion programs in parallel, even in the presence of dynamically partitioned and aliased data.  We have
also given a novel compositional parallel semantics, permitting a precise treatment of relaxed coherence modes;
in particular we have shown the Legion design is sound even with relaxed coherence.

### 2012-11-04 doc/types-pldi2013/evaluation.tex
spell checking corrections and space saving measures

\subsection{Checking Overhead}
\label{subsec:overhead}
The first Legion implementation consisted of a C++ library of Legion primitives
\cite{Legion12} with no checking of region memory accesses.  
When testing applications on this system we frequently encountered memory corruption due to
illegal region accesses caused by application bugs.
In many cases, this corruption occurred between nodes in the cluster or on 
GPUs, environments for which debugging tools are primitive at best.
%  Finding these bugs required either multiple {\tt gdb} sessions 
%for observing execution on different nodes simultaneously or using {\tt printf} for debugging on
%GPUs.  
To locate the application bugs causing these illegal accesses, we initially added 
dynamic checks on  all region accesses for both CPUs and GPUs which  
added considerable runtime overhead.  
To preserve the benefit of checking every access, but eliminate the cost of dynamic checks we implemented the type, privilege, and coherence checker we have described. 
We then rewrote the applications in this language and type checked them, at which point the
dynamic region access checks could be safely elided.

>>>>

\subsection{Checking Overhead}
\label{subsec:overhead}
The first Legion implementation consisted of a C++ library of Legion primitives
\cite{Legion12} with no checking of region memory accesses.  
When testing applications on this system we frequently encountered memory corruption due to
illegal region accesses caused by application bugs.
In many cases, this corruption occurred between nodes in the cluster or on 
GPUs, environments for which debugging tools are primitive at best.
%  Finding these bugs required either multiple {\tt gdb} sessions 
%for observing execution on different nodes simultaneously or using {\tt printf} for debugging on
%GPUs.  
To locate the application bugs causing these illegal accesses, we initially added 
dynamic checks on  all region accesses for both CPUs and GPUs which  
added considerable runtime overhead.  
To preserve the benefit of checking every access without the cost of dynamic checks, we implemented the type, privilege, and coherence checker we have described. 
We then rewrote the applications in this language and type checked them, at which point the
dynamic region access checks could be safely elided.

### 2012-11-04 doc/types-pldi2013/example.tex
spell checking corrections and space saving measures

Partitioning of the graph is done using {\em colorings} built in the
{\tt color\_circuit} function (line 18-19).  A coloring maps region
elements to colors (Core Legion uses integers), with each color corresponding
 to a new subregion
to be created.  Note that it is impossible to pick a single static
partitioning scheme that would do well for all graphs.  Dynamic
partitioning using colorings provides the programmer the flexibility
to make partitioning decisions based on input data.

@@

After creating the colorings, the application partitions the node and wire
regions into {\em subregions} according to the colorings (lines 21-25).  For
simplicity, our example only partitions the circuit into two pieces.
Line 21 uses coloring {\tt pc.1} to partition {\tt rn} into {\tt rn0}, and {\tt rn1}.
Subregion {\tt rn0} will contain all locations with a color of 0 in {\tt pc.1},
while {\tt rn1} will contain all locations with a color of 1.  This
partitioning uses a coloring and results in {\em disjoint} subpartitions.
Constraints are introduced into the static environment describing both the
disjointness of the subregions (i.e. {\tt rn0 $*$ rn1}) and the inclusion of
each subregion in the original region (e.g. {\tt rn0 $\leq$ rn}).  Line
23 is similar, but uses a multicoloring, so the subregions may be {\em aliased},
and only the inclusion constraints are introduced into the static environment.

@@

After partitioning the circuit into pieces, the application creates
instances of {\tt CircuitPiece} (defined on lines 9-12) for each piece
(lines 28-33).  {\tt CircuitPiece} is a {\em region relationship}, a
bounded existential type.  Region relationships allow the programmer
to {\em pack} a group of regions and pointers into those regions together
and remember properties
about them such as disjointness and subregion relationships (lines 29
and 33).  The type system verifies the properties hold statically when
packing; when region relationships are {\em unpacked} the same
properties are reintroduced into the static checking environment
(lines 40-41).  A key feature underlying the soundness of the Legion
type system is that privileges cannot be packed in a region
relationship---privileges belong to functions. When a function unpacks
a region relationship it must already hold privileges for the new
regions it finds in the packed value.  For example, on line 40 the
function {\tt execute\_time\_steps} has read/write privileges for
region {\tt rn0} because it has read/write privileges for {\tt rn} and
the {\tt CircuitPiece} region relationship constraints ensure $\tt rn0
\leq rn$ (line 12).

>>>>

Partitioning of the graph is done using {\em colorings} built in the
{\tt color\_circuit} function (line 18-19).  A coloring maps region
elements to colors (integers in Core Legion), with each color corresponding
 to a new subregion
to be created.  Note that it is impossible to pick a single static
partitioning scheme that would do well for all graphs.  Dynamic
partitioning using colorings enables the programmer 
to make partitioning decisions based on input data.

@@

After creating the colorings, the application partitions the node and wire
regions into {\em subregions} according to the colorings (lines 21-25).  For
simplicity, our example only partitions the circuit into two pieces.
Line 21 uses coloring {\tt pc.1} to partition {\tt rn} into {\tt rn0}, and {\tt rn1}.
Subregion {\tt rn0} will contain all locations with a color of 0 in {\tt pc.1},
while {\tt rn1} will contain all locations with a color of 1.  This
partitioning results in {\em disjoint} subpartitions.
Constraints are introduced into the static environment describing both the
disjointness of the subregions (i.e. {\tt rn0 $*$ rn1}) and the inclusion of
each subregion in the original region (e.g. {\tt rn0 $\leq$ rn}).  Line
23 is similar, but uses a multicoloring, so the subregions may be {\em aliased},
and only the inclusion constraints are introduced into the static environment.

@@

After partitioning the circuit into pieces, the application creates
instances of {\tt CircuitPiece} (defined on lines 9-12) for each piece
(lines 28-33).  {\tt CircuitPiece} is a {\em region relationship}, a
bounded existential type.  Region relationships allow the programmer
to {\em pack} a group of regions and pointers into those regions together
and remember properties
about them such as disjointness and subregion relationships (lines 29
and 33).  The type system verifies the properties hold statically when
packing; when region relationships are {\em unpacked} the same
properties are reintroduced into the static checking environment
(lines 40-41).  A key feature underlying the soundness of the Legion
type system is that privileges cannot be packed in a region
relationship---privileges belong to functions. When a function unpacks
a region relationship it must already hold privileges for the unpacked regions. 
%new regions it finds in the packed value.  
For example, on line 40 the
function {\tt execute\_time\_steps} has read/write privileges for
region {\tt rn0} because it has read/write privileges for {\tt rn} and
the {\tt CircuitPiece} region relationship constraints ensure $\tt rn0
\leq rn$ (line 12).

### 2012-11-04 doc/types-pldi2013/intro.tex
spell checking corrections and space saving measures

Previous work has explored two approaches for describing the organization of data
in distributed memory programs.
One class of work uses fully static analyses with no runtime overhead, but 
disallows potential data aliasing to make the analyis tractable, limiting expressivity.
Two recent examples, Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap.
% into a tree of collections of data.
The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be checked statically 
(see Section~\ref{sec:related}).  The second approach permits potential data
aliasing, allowing greater expressivity, but uses fully dynamic analyses of
program execution to detect data aliasing and guarantee correct results.
This approach incurs significant runtime overhead and requires 
centralized control logic that is difficult to scale to distributed 
memory machines; examples include transactional memory \cite{Harris05} 
and thread-level speculation \cite{Steffan00}.

@@

To support additional parallelism in the presence of aliased data,
Legion provides {\em coherence} modes at the granularity of regions,
which allow relaxed constraints on task execution order, even
in case where tasks do access the same region.  The programmer may
specify that tasks conflicting on a region must execute serially (in
program order, the default), {\em atomically} with respect to each
other, or even {\em simultaneously} if the application has alternative
methods for ensuring determinism or can tolerate non-deterministic
behavior in that region.

%To guarantee the safety and correctness of programs, Legion relies on
%a combination of static and dynamic analysis based on the privilege
%and coherence systems.  
%To efficiently guarantee the safety and
%correctness of programs using first-class regions,
%privilege and coherence properties in the presence of arbitrary region
%aliasing, Legion relies on a combination of static and
%dynamic checks.  
The critical feature in Legion that makes static analysis tractable
and dynamic analysis efficient in the presence of potential aliasing is logical regions.
Alias analysis is a very hard static analysis problem, but is both easy and
inexpensive if done dynamically at the granulatiry of logical regions instead
of individual memory locations; similarly
checks on individual memory accesses are expensive if done dynamically, but
cheap if done statically within the context of local logical region privileges.
%A key theme that runs through the design is that 
%alias analysis is a very hard static analysis problem, but is both easy and
%inexpensive if done dynamically at the granularity of regions instead of individual
%heap locations, while checks that must be done on individual values are only
%cheap if done statically.  
Explicating and evaluating the design of Legion's
static and dynamic semantics and its implications are the topics of this paper,
which we present as a series of contributions:
\begin{itemize}
\item We present a type system for {\em Core Legion} programs that statically
  verifies the safety of individual pointers and region
  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
%  A subtle issue arises in checking privileges when regions are
%  stored in and later retrieved from heap data structures.  The type
%  system provides enough static information that the Legion runtime
%  can leverage its dynamic information about region aliasing to
%  perform dynamic privilege checks guaranteeing safety.

>>>>

Previous work has explored two approaches for describing the organization of data
in distributed memory programs.
One class of work uses fully static analyses with no runtime overhead, but 
disallows potential data aliasing to make the analysis tractable, limiting expressivity.
Two recent examples, Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap.
% into a tree of collections of data.
The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be checked statically 
(see Section~\ref{sec:related}).  The second approach permits potential data
aliasing, allowing greater expressivity, but uses fully dynamic analyses of
program execution to detect data aliasing and guarantee correct results.
This approach incurs significant runtime overhead and requires 
centralized control logic that is difficult to scale to distributed 
memory machines; examples include transactional memory \cite{Harris05} 
and thread-level speculation \cite{Steffan00}.

@@

To support additional parallelism in the presence of aliased data,
Legion provides {\em coherence} modes at the granularity of regions,
which allow relaxed constraints on task execution order, even
in cases where tasks access the same region.  The programmer may
specify that tasks conflicting on a region run: in
program order (the default), {\em atomically} with respect to each
other, or even {\em simultaneously} if the application has alternative
methods for ensuring determinism or can tolerate non-deterministic
behavior in that region.

%To guarantee the safety and correctness of programs, Legion relies on
%a combination of static and dynamic analysis based on the privilege
%and coherence systems.  
%To efficiently guarantee the safety and
%correctness of programs using first-class regions,
%privilege and coherence properties in the presence of arbitrary region
%aliasing, Legion relies on a combination of static and
%dynamic checks.  
The critical feature in Legion that makes static analysis tractable
and dynamic analysis efficient in the presence of potential aliasing is logical regions.
Alias analysis is a very hard static analysis problem, but is both easy and
inexpensive if done dynamically at the granularity of logical regions instead
of individual memory locations; similarly
checks on individual memory accesses are expensive if done dynamically, but
cheap if done statically within the context of local logical region privileges.
%A key theme that runs through the design is that 
%alias analysis is a very hard static analysis problem, but is both easy and
%inexpensive if done dynamically at the granularity of regions instead of individual
%heap locations, while checks that must be done on individual values are only
%cheap if done statically.  
Explicating and evaluating the design of Legion's
static and dynamic semantics is the topic of this paper, 
%and its implications are the topics of this paper,
which we present as a series of contributions:
\begin{itemize}
\item We present a type system for {\em Core Legion} programs that statically
  verifies the safety of individual pointers and region
  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
%  A subtle issue arises in checking privileges when regions are
%  stored in and later retrieved from heap data structures.  The type
%  system provides enough static information that the Legion runtime
%  can leverage its dynamic information about region aliasing to
%  perform dynamic privilege checks guaranteeing safety.

### 2012-11-04 doc/types-pldi2013/langdef.tex
spell checking corrections and space saving measures

The {\em pack} and {\em unpack} expressions allow regions and pointers into those regions to be stored in the heap.
%  Logical region names are lexically scoped and if stored in the heap as-is, can only be read within the same scope.  
A {\em region relationship} captures statically known constraints between regions;
because the {\em pack} rule verifies these constraints hold when the 
region relationship value is created,{\em unpack} can reintroduce these constraints on
the (fresh) logical region names when the value is read and unpacked from the heap.
Crucially for soundness,  region relationships do not capture privileges.
%
% FIXME  It would be nice to add some comments about unpack here, and how the constraints introduced by unpack
% must be sufficient to allow access to the new regions if they are to be used, which generally requires the type
% have a free region variable capturing subregioning relationships between the packed regions and some existing region
% in scope for which privileges are held.
% 
Finally, task calls verify the requirement that the privileges held by a called task are a subset of those held
by the caller.  In addition to being necessary for the proof of soundness, this property is critical to
enabling hierarchical and distributed scheduling of tasks.

@@

Instead we use a bottom-up approach with big-step semantics.
Each expression has a clobber set $C$ of memory locations that might 
be modified by concurrent evaluations.  While a read from a location not in $C$ is {\em coherent}
(returning the most recently written value), the value read from a location $l$ in $C$ may be any
value.  Determing which value is deferred until all potentially-concurrent
modifications to $l$ are observable in the evaluation of a parent expression whose clobber set does
not include $l$ and a consistent interleaving of the subexpressions' memory traces is determined.
%With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
%made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithm
%used by the Legion runtime satisfies these constraints, guaranteeing the soundness of the Legion programming model.

@@

The proof, which can be found in the submitted appendix, is made
straightforward by the compositional operational semantics.
%It proceeds by standard induction on the structure of the
%derivation in the operational semantics, at each step showing that
%subexpressions have consistent starting environments and that the
%value and trace consistency of the subexpressions is sufficient to prove
%the consistency of the parent expression.

@@

\begin{thm}
\label{thm:atomicexec}
\rm
Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
Let $M$ be a region mapping, $L$ a local value mapping,
$H$ a heap typing, and $S$ be an initial store satisfying
$\mapconsist{\Omega}$, $\localconsist{L}{\Gamma}$, and $\storeconsist{S}{H}$.
If $\Phi_i \nonint[M]^A \Phi_j$ for $1 \leq i < j \leq n$, then for any permuatation $(\oton{\pi})$ of
$(1, \ldots, n)$, $E_{\pi_1} \dplus \ldots \dplus E_{\pi_n}$ is a valid interleaving.
\end{thm}

>>>>

The {\em pack} and {\em unpack} expressions allow regions and pointers into those regions to be stored in the heap.
%  Logical region names are lexically scoped and if stored in the heap as-is, can only be read within the same scope.  
A {\em region relationship} captures statically known constraints between regions;
because the {\em pack} rule verifies these constraints hold when the 
region relationship value is created, {\em unpack} can reintroduce these constraints on
the (fresh) logical region names when the value is unpacked from the heap.
Crucially for soundness, region relationships do not capture privileges.
%
% FIXME  It would be nice to add some comments about unpack here, and how the constraints introduced by unpack
% must be sufficient to allow access to the new regions if they are to be used, which generally requires the type
% have a free region variable capturing subregioning relationships between the packed regions and some existing region
% in scope for which privileges are held.
% 
Task calls verify the requirement that the privileges held by a called task are a subset of those held
by the caller.  In addition to being necessary for the proof of soundness, this property is critical for 
hierarchical and distributed scheduling of tasks.

@@

Instead we use a bottom-up approach with big-step semantics.
Each expression has a clobber set $C$ of memory locations that might 
be modified by concurrent evaluations.  While a read from a location not in $C$ is {\em coherent}
(returning the most recently written value), the value read from a location $l$ in $C$ may be any
value.  Determining which value is deferred until all potentially-concurrent
modifications to $l$ are observable in the evaluation of a parent expression whose clobber set does
not include $l$ and a consistent interleaving of the subexpressions' memory traces is determined.
%With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
%made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithm
%used by the Legion runtime satisfies these constraints, guaranteeing the soundness of the Legion programming model.

@@

The proof is made straightforward by the compositional operational
semantics.  The proof, along with proofs of all theorems and lemmas
in this paper, can be found in the submitted appendix.
%The proof, which can be found in the submitted appendix, is made
%straightforward by the compositional operational semantics.
%It proceeds by standard induction on the structure of the
%derivation in the operational semantics, at each step showing that
%subexpressions have consistent starting environments and that the
%value and trace consistency of the subexpressions is sufficient to prove
%the consistency of the parent expression.

@@

\begin{thm}
\label{thm:atomicexec}
\rm
Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
Let $M$ be a region mapping, $L$ a local value mapping,
$H$ a heap typing, and $S$ be an initial store satisfying
$\mapconsist{\Omega}$, $\localconsist{L}{\Gamma}$, and $\storeconsist{S}{H}$.
If $\Phi_i \nonint[M]^A \Phi_j$ for $1 \leq i < j \leq n$, then for any permutation $(\oton{\pi})$ of
$(1, \ldots, n)$, $E_{\pi_1} \dplus \ldots \dplus E_{\pi_n}$ is a valid interleaving.
\end{thm}
